---
title: "Boston Marathon Loess Example"
author: "rbtik"
---
In this assignment, we are tasked with scraping Boston Marathon run times from Wikipedia in order to plot the results.

First step is to load the necessary libraries.

```{r}
library(XML)
library(RCurl)
library(lubridate)
library(ggplot2)
```

In this next code we 1) parse the HTML into R Data Frames, 2) combine genders as they were originally separated into two data tables, 3) transform the time it took to run the marathon into seconds.

```{r}
URL <- 'http://en.wikipedia.org/wiki/List_of_winners_of_the_Boston_Marathon'
tables <- readHTMLTable(URL)

males <- tables[[5]]
females <- tables[[6]]
males$gender <- 'Male'
females$gender <- 'Female'
df <- rbind(males, females)

to_seconds <- function(t) {
  time <- hms(t)
  return(hour(time) * 3600 + minute(time) * 60 + second(time))
}
df$timeInSec <- to_seconds(df$Time)
```

Finally we plot the DataFrames and apply the Loess method.

```{r}
ggplot(df, aes(x=Year, y=timeInSec, group=gender, col=gender, fill=gender)) +
      geom_point() +
      geom_smooth(size=1, method="loess") + 
      xlab("Years")
```

